---
title: "Portfolio"
author: "Niels Rouws, 12177555"
date: "2/10/2021"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}

#load("Env.RData")

library(tidyverse)
library(spotifyr)
library(flexdashboard)
library(plotly)
library(compmus)
library(gridExtra)

```

```{r, echo=FALSE, cache=TRUE}
rock60s <- get_playlist_audio_features("", "37i9dQZF1DWWzBc3TOlaAV") %>% add_audio_analysis()
rock70s <- get_playlist_audio_features("", "37i9dQZF1DWWwzidNQX6jx") %>% add_audio_analysis()
rock80s <- get_playlist_audio_features("", "37i9dQZF1DX1spT6G94GFC") %>% add_audio_analysis()
rock90s <- get_playlist_audio_features("", "37i9dQZF1DX1rVvRgjX59F") %>% add_audio_analysis()
rock00s <- get_playlist_audio_features("", "37i9dQZF1DX3oM43CtKnRV") %>% add_audio_analysis()
rock10s <- get_playlist_audio_features("", "37i9dQZF1DX99DRG9N39X3") %>% add_audio_analysis()


rock <-
  bind_rows(
    rock60s %>% mutate(category = "1960s"),
    rock70s %>% mutate(category = "1970s"),
    rock80s %>% mutate(category = "1980s"),
    rock90s %>% mutate(category = "1990s"),
    rock00s %>% mutate(category = "2000s"),
    rock10s %>% mutate(category = "2010s")
  ) %>%
  mutate(
    mode = ifelse(mode == 0, "Minor", "Major"),
    track.album.release_date = as.numeric(substr(track.album.release_date, 1, 4))
  )

```

Introduction {.storyboard}
=========================================

### Comparing the Rock genre over four decades 60s-10s

Spotify has 6 playlist called the "XXs Rock Anthems" for the decades 60s, 70s, 80s, 90s, 00s, and 10s. I'm interested
in how rock has evolved from the 60s to the 10s. Multiple bands have made music spanning more than one decade
and I would like to investigate how their earlier work compares to their later works and whether bands have
influenced each other. Moreover when large subgenres of rock started to appear and which bands were responsible. 
There are three main groups to compare: the early bands from the 60s and 70s, the middle 
80 and 90s bands, and late 00s and 10s. So the main questions i want to answer is:

* What are the most notable changes that happened in the rock genre from the 60s to the 10s?
	
The use of more digital elements in music will probably increase over the years, and more subgenres are likely to have
emerged over the years. I am unsure in what way the bands have influenced each other if any at all.

The tracks are all popular songs from their respective time periods but not all songs are original. 
Especially among the older songs are a lot of remastered versions. Another downside is the small 
amount of songs representing a complete decade of song releases. 

Atypical bands might be bands like Queen which is known for releasing songs in a multitude of genres 
and bands that have released albums spanning multiple decades. Typical bands might be bands like U2 that appear 
in one playlist with multiple songs.

The playlists will be analyzed on multiple levels. The decades as a collective shall be used to show trends and 
the general characteristics of the playlists. For a more detailed comparison only look the most popular
songs per decade or special cases will be analyzed. 

The two most popular songs per decade according to the Spotify API are:

* 1960s
  - [Fortunate Son - Creedance Clearwater Revival](https://open.spotify.com/track/4BP3uh0hFLFRb5cjsgLqDh?si=uxgBjQ8eTFCmeK5u7gNHrA)
  - [Brown Eyed Girl - Van Morrison](https://open.spotify.com/track/3yrSvpt2l1xhsV9Em88Pul?si=8Z8i1r_WSJ6QBmYICTwGMA)
* 1970s
  - [Hotel California - 2013 Remaster](https://open.spotify.com/track/3yrSvpt2l1xhsV9Em88Pul?si=8Z8i1r_WSJ6QBmYICTwGMA)
  - [Rocket Man (I Think It's Going To Be A Long, Long Time) - Elton John](https://open.spotify.com/track/3gdewACMIVMEWVbyb8O9sY?si=aR6wxMGzQ_SRt8kersqdpA)
* 1980s
  - [Back in Black - AC/DC](https://open.spotify.com/track/08mG3Y1vljYA6bvDt4Wqkj?si=dFnbaIOVR1SA6fvYLjGsqQ)
  - [Take on me - a-ha](https://open.spotify.com/track/2WfaOiMkCvy7F5fcp2zZ8L?si=IlPDcyUjScqSs1eigatM_g)
* 1990s
  - [Zombie - The Cranberries](https://open.spotify.com/track/2IZZqH4K02UIYg5EohpNHF?si=MyMspx6PQTWi3SbHJ-CYZQ)
  - [Under the Bridge - Red Hot Chili Peppers](https://open.spotify.com/track/3d9DChrdc6BOeFsbrZ3Is0?si=tCS9HeYPTuCd6mHIlhmb4g)
* 2000s
  - [Can't Stop - Red Hot Chili Peppers](https://open.spotify.com/track/3ZOEytgrvLwQaqXreDs2Jx?si=i2mQLeJ3RPqu_dZ1f59Y3Q)
  - [You're Gonna Go Far, Kid - The Offspring](https://open.spotify.com/track/6TfBA04WJ3X1d1wXhaCFVT?si=rBeCnIp8Sliv77ndT5_nNg)
* 2010s
  - [Believer - Imagine Dragons](https://open.spotify.com/track/0pqnGHJpmpxLKifKRmU6WP?si=Lw12GL48SH6Ysu_zauCqvA)
  - [Do I Wanna Know? - Artic Monkeys](https://open.spotify.com/track/5FVd6KXrgO9B3JPmC8OPst?si=XcQhJ91TTSqBa3SzM-yc0A)



The Playlists {.storyboard}
=========================================

### How many songs are in the "XXs Rock Anthems"?

```{r, echo=FALSE}

ggplot(
  rock,
  aes(
    x = category,
    fill = category
  )) +
    geom_bar() +
    labs(
    title = "Amount of songs per decennium",
    x = "decennium",
    y = "song count"
  ) +
  theme_light() +
  theme(legend.position = "none")

```

***

The current corpus has a lack of data representing the last two decades and the 
90s are over-represented. Furthermore the current data might not be the best 
to represent the decennia since they have not yet been verified using the 
hit-charts of their respective decade.



### Release dates and popularity

```{r, echo=FALSE}

plot_pop <- ggplot(
  rock,
  aes(
    x = track.album.release_date,
    y = track.popularity,
    color = category
  )) +
    geom_point( size = 2, alpha = 0.6) +
    labs(
    title = "Release date and popularity of songs",
    x = "Release date",
    y = "Popularity",
    legend.title = "Rock Anthem playlist"
  ) +
  theme_light() +
  scale_x_continuous(breaks = c(1960,1970,1980,1990,2000,2010))

ggplotly(plot_pop, tooltip=c("track.name", "x", "y"))

```


***

The corpus contains several remasters which can been seen as dots that don't bear
the color of the decade they are in. Clear examples can be seen on the year 2000
where several 1960s, the orange colored ones, songs are present. This explains dots
showing up later than their category but two 1980s songs have been released in the 
before the start of their assigned decade. The popular [Barracuda - Heart](https://open.spotify.com/track/4KfSdst7rW39C0sfhArdrz?si=shk5j6fCSNeW8KAsEOJV9Q)
was released in 1977 and [I Was Made For Lovin' You - KISS](https://open.spotify.com/track/2ZSCy3P1QzpJySCWA6NRIU?si=LlsrX48GToifahxfcwVsbA) in 1979.

It's quite remarkable that there is a handful of songs each decade that have low popularity scores which they might have lost over the years but does give
us insight that current popularity isn't the only factor Spotify have looked at making these playlists.



Track-level Features {.storyboard}
=========================================


### The acousticness decreased significantly over the decades


```{r, echo=FALSE, cache=TRUE}

ggplot(
  rock,
  aes(
    x = category,
    y = acousticness,
    fill = mode
  )
) +
  geom_boxplot(
    position = position_dodge(width = 0.5),
    outlier.alpha = 0.6,
  ) +
  geom_text(aes(label=ifelse(track.name=="Way down We Go", "Way down We Go", "")), hjust = 0.5, vjust=-0.5) +
  theme(legend.position = c(0.9, 0.9)) +
  labs(title = " Distribution of acousticness per decennia per mode")

```

***

The current data shows that over the years the acousticness has significantly 
decreased. From the 1980s a significant drop in acousticness can be seen in 
the data. The 1960s and 1970s show a wide distribution which narrows and shifts
towards no acousticness at all. The 2000s seems the decade with the least
acousticness as even the outliers don't exceed 0.12 on the scale. The 2010s see
the acoustic outliers returning with "Way down We Go" from KALEO as most
acoustic song.


The distribution along the modes don't seem to have a significant preference for
more or less acousticness only the 1960s and 1980s show dissimilar
distributions per mode. Minor songs favor more acousticsness in the 1960s and
the 1980s is reversed, favoring less acousticness.


### The early rock years were more valent too



```{r, echo=FALSE, cache=TRUE}
ggplot(
  rock,
  aes(
    x = category,
    y = valence,
    fill = mode
  )
) +
  geom_boxplot(
    position = position_dodge(width = 0.5),
    outlier.alpha = 0.6,
  ) +
  theme(legend.position = c(0.9, 0.9)) +
  labs(title = "Distribution of valence per decennia per mode")
```

***

The valence appears to barely have changed in the first three decades with a 
small drop to the last three decades.

### Rock has become more energetic and louder


```{r, echo=FALSE, cache=TRUE}
plot_le1 <- ggplot(
  rock,
  aes(
    x = loudness,
    y = energy,
    color = category,
    name = track.name,
    size = danceability
  )
) +
  geom_point(alpha = 0.4) +
  facet_wrap(~category) +
  theme(legend.position = "none") +
  labs(
    title = "Energy related to Loudness (dB) per decennium.",
    x = "Loudness (dB)",
    y = "Energy"
  ) +
  theme_light()





ggplotly(plot_le1, tooltip=c("track.name", "x", "y", "size"))


```

***

Comparing the energy related to loudness per decade show very similar clouds of
data points. Over the decades the representation in the low loudness and energy seems 
to have decreased suggesting that songs have become louder and more energetic. 

It should be kept in mind that the 00s and 10s playlist have the lowest amount of 
songs. The 1990s is the largest playlist however which has sparsity in the low energy
and loudness region as well compared to the 60s and 70s so the trend seems to be that 
songs have high energy and loudness. 



### Danceability has not seen any big changes

```{r, echo=FALSE, cache=TRUE}
plot_le2 <- ggplot(
  rock,
  aes(
    x = loudness,
    y = danceability,
    size = energy,
    color = category,
    name = track.name,
  )
) +
  geom_point(alpha = 0.4) +
  facet_wrap(~category) +
  theme(legend.position = "none") +
  labs(
    title = "Danceability related to Loudness (dB) per decennium.",
    x = "Loudness (dB)",
    y = "Danceability"
  ) +
  theme_light()
ggplotly(plot_le2, tooltip=c("track.name", "x", "y", "size"))

```


***

The distribution of danceability has not seen any huge changes over the decades. It's quite notable that 
a wide range of dancebility is present in the playlists. The values range from low 0.25 songs to high 0.75 songs.
with a few outliers such as the 80s song [Another One Bites The Dust - Remastered 2011 - Queen](https://open.spotify.com/track/57JVGBtBLCfHw2muk5416J?si=CxzaiY22QNOehHyo2EMqpA)



Chroma features {.storyboard}
=========================================

### Fortunate Son versus Believer


```{r, echo=FALSE, cache=TRUE}

ssm1 <-
  get_tidy_audio_analysis("4BP3uh0hFLFRb5cjsgLqDh") %>%
  compmus_align(bars, segments) %>%
  select(bars) %>%
  unnest(bars) %>%
  mutate(
    pitches =
      map(segments,
        compmus_summarise, pitches,
        method = "acentre", norm = "manhattan"
      )
  ) %>%
  mutate(
    timbre =
      map(segments,
        compmus_summarise, timbre,
        method = "mean"
      )
)

plot_ssm1 <- bind_rows(
  ssm1 %>%
    compmus_self_similarity(pitches, "aitchison") %>%
    mutate(d = d / max(d), type = "Chroma"),
  ssm1 %>%
    compmus_self_similarity(timbre, "euclidean") %>%
    mutate(d = d / max(d), type = "Timbre")
) %>%
  mutate() %>%
  ggplot(
    aes(
      x = xstart + xduration,
      width = xduration,
      y = ystart + yduration,
      height = yduration,
      fill = d
    )
  ) +
  geom_tile() +
  coord_fixed() +
  facet_wrap(~type) +
  scale_fill_viridis_c(guide = "none") +
  theme_light() +
  labs(title = "Fortunate Son - Creedance Clearwater Revival", x = "Time (s)", y = "Time (s)")

##########################

ssm2 <-
  get_tidy_audio_analysis("0pqnGHJpmpxLKifKRmU6WP") %>%
  compmus_align(bars, segments) %>%
  select(bars) %>%
  unnest(bars) %>%
  mutate(
    pitches =
      map(segments,
        compmus_summarise, pitches,
        method = "acentre", norm = "manhattan"
      )
  ) %>%
  mutate(
    timbre =
      map(segments,
        compmus_summarise, timbre,
        method = "mean"
      )
  )


plot_ssm2 <- bind_rows(
  ssm2 %>%
    compmus_self_similarity(pitches, "aitchison") %>%
    mutate(d = d / max(d), type = "Chroma"),
  ssm2 %>%
    compmus_self_similarity(timbre, "euclidean") %>%
    mutate(d = d / max(d), type = "Timbre")
) %>%
  mutate() %>%
  ggplot(
    aes(
      x = xstart + xduration,
      width = xduration,
      y = ystart + yduration,
      height = yduration,
      fill = d
    )
  ) +
  geom_tile() +
  coord_fixed() +
  facet_wrap(~type) +
  scale_fill_viridis_c(guide = "none") +
  theme_light() +
  labs(title = "Believer - Image Dragons", x = "Time (s)", y = "Time (s)")


grid.arrange(plot_ssm1, plot_ssm2)

```


***





### Hotel California - Original versus Remaster

```{r, echo=FALSE, cache=TRUE}

hc_original <-
  get_tidy_audio_analysis("1rh232CwAy3EDEWFJkwH88") %>%
  select(segments) %>%
  unnest(segments) %>%
  select(start, duration, pitches)

hc_remaster <-
  get_tidy_audio_analysis("40riOy7x9W7GXjyGp4pjAv") %>%
  select(segments) %>%
  unnest(segments) %>%
  select(start, duration, pitches)


plot_chroma <- hc_remaster %>%
  mutate(pitches = map(pitches, compmus_normalise, "euclidean")) %>%
  compmus_gather_chroma() %>%
  ggplot(
    aes(
      x = start + duration / 2,
      width = duration,
      y = pitch_class,
      fill = value
    )
  ) +
  geom_tile() +
  labs(title="Chromagram Hotel California - 2013 Remaster", x = "Time (s)", y = NULL, fill = "Magnitude") +
  theme_minimal() +
  scale_fill_viridis_c()



plot_dwt <- compmus_long_distance(
  hc_original %>% mutate(pitches = map(pitches, compmus_normalise, "chebyshev")),
  hc_remaster %>% mutate(pitches = map(pitches, compmus_normalise, "chebyshev")),
  feature = pitches,
  method = "euclidean"
) %>%
  ggplot(
    aes(
      x = xstart + xduration / 2,
      width = xduration,
      y = ystart + yduration / 2,
      height = yduration,
      fill = d
    )
  ) +
  geom_tile() +
  coord_equal() +
  labs(x = "1976 Original", y = "2013 Remaster") +
  theme_minimal() +
  scale_fill_viridis_c(guide = NULL)


grid.arrange(plot_chroma, plot_dwt)

```


***

[Hotel California 1976](https://open.spotify.com/track/1rh232CwAy3EDEWFJkwH88?si=L6fxlcitSMGN6NQr5kAeog)
[Hotel California - 2013 Remaster](https://open.spotify.com/track/3yrSvpt2l1xhsV9Em88Pul?si=8Z8i1r_WSJ6QBmYICTwGMA)


Tempo and Timbre {.storyboard}
=========================================

### Tempo distribution


```{r, echo=FALSE, cache=TRUE}

# rock %>%
#   mutate(
#     sections =
#       map(
#         sections,                                    # sections or segments
#         summarise_at,
#         vars(tempo, loudness, duration),             # features of interest
#         list(section_mean = mean, section_sd = sd)   # aggregation functions
#       )
#   ) %>%
#   unnest(sections) %>%
#   ggplot(
#     aes(
#       x = tempo,
#       y = tempo_section_sd,
#       colour = category,
#       alpha = loudness
#     )
#   ) +
#   geom_point(aes(size = duration / 60)) +
#   geom_rug() +
#   facet_wrap(~category) +
#   theme_minimal() +
#   ylim(0, 5) +
#   labs(
#     x = "Mean Tempo (bpm)",
#     y = "SD Tempo",
#     colour = "Decade",
#     size = "Duration (min)",
#     alpha = "Volume (dBFS)"
#   )

```

***

Lorem ipsum


### Distribution of timbre coefficients

```{r, echo=FALSE, cache=TRUE}

rock %>%
  mutate(
    timbre =
      map(
        segments,
        compmus_summarise,
        timbre,
        method = "mean"
      )
  ) %>%
  select(category, timbre) %>%
  compmus_gather_timbre() %>%
  ggplot(aes(x = basis, y = value, fill = category)) +
  geom_violin() +
  scale_fill_viridis_d() +
  labs(x = "Spotify Timbre Coefficients", y = "", fill = "Decennium") +
  theme_light() +
  theme(legend.position = "none")

```

***

The timbre coefficients for each decade are closely related. All decades have distributions near the 
same region for each coefficient. Finding features that can reliably split decades are not present. A extremely 
tiny trends can be seen in most coefficients as for example in the first or sixth coefficients but other than
those examples are the timbre coefficients not great features to distinguish the playlist based on them


Tempo analysis of two iconic 70s songs {.storyboard}
=========================================

### Led Zeppelin - Stairway to Heaven & Queen - Bohemian Rhapsody


```{r, echo=FALSE, cache=TRUE}

zeppelin <- get_tidy_audio_analysis("5CQ30WqJwcep0pYcV4AMNc")

plot_zep <-zeppelin %>%
  tempogram(window_size = 8, hop_size = 1, cyclic = FALSE) %>%
  ggplot(aes(x = time, y = bpm, fill = power)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  labs(title = "Tempogram Led Zeppelin - Stairway to Heaven", x = "Time (s)", y = "Tempo (BPM)") +
  theme_light()

rhapsody <- get_tidy_audio_analysis("2GAIycsMaDVtMtdvxzR2xI")

```



```{r, echo=FALSE, cache=TRUE}
plot_rhap <- rhapsody %>%
  tempogram(window_size = 8, hop_size = 1, cyclic = FALSE) %>%
  ggplot(aes(x = time, y = bpm, fill = power)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  labs(title = "Tempogram Queen - Bohemian Rhapsody", x = "Time (s)", y = "Tempo (BPM)") +
  theme_light()



grid.arrange(plot_zep, plot_rhap)
```


***

Two iconic rock numbers have interesting tempi changes. Stairway to Heaven has 
a constantly increasing tempo. Starting around 71 BPM the tempo bumps up to 
around 82 BPM and continues to increase. At the end of verse 6, or 330 seconds,
a transition and solo follow that have a peak BPM of just over 100 BPM. 

Bohemian Rhapsody has iconic sections of different genres in the same song.
The tempo change that stands out the most is the operatic section with a BPM of 
146 starting around 190 seconds and ending around 240 seconds.



Classification & Clustering {.storyboard}
=========================================


### Random Forest

```{r, echo=FALSE, cache=TRUE}
# 
# get_conf_mat <- function(fit) {
#   outcome <- .get_tune_outcome_names(fit)
#   fit %>% 
#     collect_predictions() %>% 
#     conf_mat(truth = outcome, estimate = .pred_class)
# }  
# 
# get_pr <- function(fit) {
#   fit %>% 
#     conf_mat_resampled() %>% 
#     group_by(Prediction) %>% mutate(precision = Freq / sum(Freq)) %>% 
#     group_by(Truth) %>% mutate(recall = Freq / sum(Freq)) %>% 
#     ungroup() %>% filter(Prediction == Truth) %>% 
#     select(class = Prediction, precision, recall)
# }  
# 
# 
# rock_features <- rock %>%  # For your portfolio, change this to the name of your corpus.
#   add_audio_analysis() %>% 
#   mutate(
#     playlist = factor(category),
#     segments = map2(segments, key, compmus_c_transpose),
#     pitches =
#       map(
#         segments,
#         compmus_summarise, pitches,
#         method = "mean", norm = "manhattan"
#       ),
#     timbre =
#       map(
#         segments,
#         compmus_summarise, timbre,
#         method = "mean",
#       )
#   ) %>%
#   mutate(pitches = map(pitches, compmus_normalise, "clr")) %>%
#   mutate_at(vars(pitches, timbre), map, bind_rows) %>%
#   unnest(cols = c(pitches, timbre))
# 
# 
# rock_recipe <-
#   recipe(
#     playlist ~
#       danceability +
#       energy +
#       loudness +
#       speechiness +
#       acousticness +
#       instrumentalness +
#       liveness +
#       valence +
#       tempo +
#       duration +
#       C + `C#|Db` + D + `D#|Eb` +
#       E + `F` + `F#|Gb` + G +
#       `G#|Ab` + A + `A#|Bb` + B +
#       c01 + c02 + c03 + c04 + c05 + c06 +
#       c07 + c08 + c09 + c10 + c11 + c12,
#     data = rock_features,          # Use the same name as the previous block.
#   ) %>%
#   step_center(all_predictors()) %>%
#   step_scale(all_predictors())      # Converts to z-scores.
#   # step_range(all_predictors())    # Sets range to [0, 1].
# 
# 
# rock_cv <- rock_features %>% vfold_cv(5)
# 
# knn_model <-
#   nearest_neighbor(neighbors = 1) %>%
#   set_mode("classification") %>% 
#   set_engine("kknn")
# rock_knn <- 
#   workflow() %>% 
#   add_recipe(rock_recipe) %>% 
#   add_model(knn_model) %>% 
#   fit_resamples(
#     rock_cv, 
#     control = control_resamples(save_pred = TRUE)
#   )
# 
# rock_knn %>% get_conf_mat() autoplot(type = "heatmap")
# 
# forest_model <-
#   rand_forest() %>%
#   set_mode("classification") %>% 
#   set_engine("ranger", importance = "impurity")
# rock_forest <- 
#   workflow() %>% 
#   add_recipe(rock_recipe) %>% 
#   add_model(forest_model) %>% 
#   fit_resamples(
#     indie_cv, 
#     control = control_resamples(save_pred = TRUE)
#   )
# 
# 
# workflow() %>% 
#   add_recipe(rock_recipe) %>% 
#   add_model(forest_model) %>% 
#   fit(indie_features) %>% 
#   pluck("fit", "fit", "fit") %>%
#   ranger::importance() %>% 
#   enframe() %>% 
#   mutate(name = fct_reorder(name, value)) %>% 
#   ggplot(aes(name, value)) + 
#   geom_col() + 
#   coord_flip() +
#   theme_minimal() +
#   labs(x = NULL, y = "Importance")

```


```{r, echo=FALSE, cache=TRUE}

# rock_features %>%
#   ggplot(aes(x = c01, y = c02, colour = playlist, size = energy)) +
#   geom_point(alpha = 0.8) +
#   scale_color_viridis_d() +
#   labs(
#     x = "Timbre Component 1",
#     y = "Timbre Component 2",
#     size = "Energy",
#     colour = "Playlist"
#  )
```


***





### Clustering


Conclusion
=========================================





